---
title: "Pension funds in Pacific Islands"
description: "An analysis of the accounts"
date: 03-07-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(ggplot2)
library(plotly)
library(magrittr)
library(tippy)
```

Last year I helped deliver a project for the IFC and members of the Pacific Island Investment Forum (PIIF). The PIIF is an association of funds across the Pacific, including provident, superannuation, sovereign and trust funds. Examples include the Vanuatu National Provident Fund (VNPF) and the Fiji National Provident Fund (FNPF). In many Pacific Island nations they represent the largest domestic asset owner and investor and collectively have over $USD 8 billion under management. The project got me interested in understanding their role further, so I did what feels natural and started looking for data. 

Also, this is the Island Economist blog so lets talk about some island institutions!

The data I eventually found was in the financial accounts. As far as I can tell no one has ever attempted to extract this data and undertake analysis. Which doesn't surprise me, the data is held in pdfs and in some cases scanned pdfs. The effort required to extract the data was substantial and mostly manual. If any of the following analysis is incorrect its likely that I mis-typed a number.

<u>
```{r}
tippy("Of the 16 PIIF members this analysis only covers 9 where I was able to compile enough data.
     
       ", tooltip = "The 9 members are: Cook Islands National Superannuation Fund (CI), Fiji National Provident Fund (FNPF), PNG - Nambawan Super, PNG - Nasfund, Solomon Islands National Provident Fund (SINPF), Samoa National Provident Fund (SNPF), Tuvalu National Provident Fund (TNPF), Tonga Retirement Fund Board (TNRBF) and Vanuatu National Provident Fund (VNPF)")

tippy_this('testing this',"fjsdkfjsdfkj")
```
</u>
The question I asked of the data was if I had $10,000 to invest into one of these institutions would I do it?

There are three ways of thinking about this problem: risk, return and fees.

<h1>Risk and return</h1>

I first wanted to understand the returns these funds achieve. I estimated annual real returns by looking at the change in net assets every year. Net assets were adjusted for net member fund movements. Individuals make contributions to these funds and they also make withdrawals (usually when they hit retirement age). If member contributions are greater than withdrawals then without an adjustment returns will look higher. I also adjusted all figures for CPI inflation of the country the fund is located in to ensure comparability.

The figure below shows the arithmetic mean annual real return and standard deviation for each of the funds. Standard deviation is a measure of volatility and by extension risk. There seems to be three general groupings. Two high performers (Cook Islands - CI and Solomon Islands - SINPF), a group in the middle and two low performers (Vanuatu - VNPF and Tonga - TNRBF). It is however worth highlighting that SINPF's performance is driven by a large increase in the value of unlisted equity in 2013. I haven't been able to establish exactly what occured in that year. 

```{r}
PlotlyData <- read.csv("C:\\Users\\Benjamin\\Desktop\\Blog\\Analyst\\PIIF\\FullReturnData.csv", header = TRUE, stringsAsFactors = FALSE)

PlotlyData$Type <- as.character(PlotlyData$Type)

#Split data into SuperFunds and PacificFunds

SuperFunds <- PlotlyData[1:25,]
PacificFunds <- PlotlyData[26:34,]


#Plot the graph
#This has a button for turning things on and off

####Risk Return graph####

plot_ly(marker = list(size = 20)) %>% 
  add_trace(data = SuperFunds, type = "scatter", x = ~SDReturns, y = ~ArithMeanReturn, 
             text = ~Institution, name = "", 
            hoverinfo = "text",
            marker = list(color = "Blue", opacity = 0.6), visible = F
            ) %>%
  add_trace(data = PacificFunds, type = "scatter", x = ~SDReturns, y = ~ArithMeanReturn, 
           text = ~Institution, name = "", alpha = 1, hoverinfo = "text", 
           marker=list(
             color = ~Color,
                       line = list(
                         color = "Black",
                         width = 1))
           ) %>%
  layout(showlegend = FALSE,
         xaxis = list(title = "Standard deviation of returns",range = c(0, 20)),
         yaxis = list(title = "Arithmetic mean real returns",range = c(-1.1, 10))) %>%
  config(displayModeBar = FALSE)  %>% 
  layout(
    updatemenus = list(
      list(
        type = "buttons",
        y = 10, 
        buttons = list(
          
          list(method = "restyle",
               args = list("visible", c(F,T)),
               label = "Pacific Funds only"),
          
          list(method = "restyle",
               args = list("visible", c(T,T)),
               label = "All Funds")))
    ))

```

The figure above shows the Pacific funds relative to each other. But what about relative to other potential investments? I collected data on Australian superfunds (the data is easily downloadable) as they are similar to Pacific funds in many ways. They hold multiple asset classes and it is mandatory to contribute to a superfund.

I did exactly the same analysis for the top 30 superfunds by net assets in 2020. If you haven't already you can click the all funds button in the figure above to populate it. 

What really surprised me was that both in terms of average real returns and volatility Pacific funds are incredibly similar to Australian superfunds. Real returns of around 2% to 4% per year can be observed for Australian superfunds and likewise for Pacific funds. It is also certainly possible for a fund with billions of dollars under management to on average barely achieve real returns across an 11 year period (looking at you Super Directions Fund). So perhaps VNPF and TNRBF aren't that unusual. 

As an aside, there is also one stand-out - Qsuper (I hold my super at Qsuper...), but then again past returns are  not indicative of future performance.

<h1>Fees</h1>

The next area I wanted to examine was fees. Fees incurred can have a substantial impact on returns. If a fund is consuming 2% of net assets annually to maintain its operations that directly reduces returns by 2% per year. If a fund is on average achieving 2-4% returns a year, then fees of just 2% feel like a huge drag.

The figure below shows the expense ratio for PIIF members in the last year of available data. In general, expense ratios are substantially below 2%. On the high side there are SINPF and VNPF while on the low side you have FNPF and SNPF. We can again look at the top 30 Australian super funds and we find expense ratios around 0.5%. Some older evidence from the [OECD](https://www.oecd.org/finance/private-pensions/41488510.pdf) found a range between 0.5% and 2.0% for institutions similar to PIIF members. So while Pacific funds have higher expense ratios than Australian super funds they don't seem too out of line with international averages. 

```{r}

ExpenseLast <- read.csv("C:\\Users\\Benjamin\\Desktop\\Blog\\Analyst\\PIIF\\ExpenseLast.csv", header = TRUE, stringsAsFactors = FALSE)

P3 <- ggplot(ExpenseLast, aes(x=reorder(Institution,ExpenseRatio), y=ExpenseRatio, fill = Institution.name)) + 
  geom_col()  + theme_bw() + theme(legend.position = "none") + scale_y_continuous(expand = c(0,0), limits = c(0,3))

P4 <- ggplotly(P3,  tooltip = c("ExpenseRatio","Institution.name")) %>% config(displayModeBar = FALSE) %>% layout(xaxis = list(title = "Institution"), yaxis = list(title = "Expense Ratio (Percent)"))

P4

```

The next figure shows the expense ratios for each of these institutions over time. FNPF has been removed as apart from the last two years of data I extracting data for the consolidated entity, which results in an uninformative measure of the expense ratio (I may go back and fix this if I find the energy).

```{r}

ExpenseOverTime <- read.csv("C:\\Users\\Benjamin\\Desktop\\Blog\\Analyst\\PIIF\\ExpenseOverTime.csv", header = TRUE, stringsAsFactors = FALSE)

P <- ggplot(ExpenseOverTime, aes(x=Year, y=ExpenseRatio, color = Institution.name)) + geom_line(size = 1.2  ) + 
  theme_bw() + facet_wrap(~Institution, nrow = 2) + theme(legend.position = "none") +
  scale_x_continuous(breaks = c(2010,2013,2016,2019)) + xlab("")

P2 <- ggplotly(P) %>% config(displayModeBar = FALSE) 
P2

```

We find that in general expense ratios are either stable or declining. There are a few anomalies in the data most visible for VNPF but also PNGNambawan. There are obvious spikes in expenses in some years and these are associated with write-downs/doubtful debts. These are recorded in the summary financial statements as expenses. Perhaps a more accurate expense ratio measure would remove these one-off impacts. However, this would require extracting the data in the notes of the financial accounts, which is a bridge too far for this blog post.

<h1>Conclusions and next time</h1>

To answer the question  I started with, would I invest into one of these funds? This preliminary analysis suggests that these funds are doing about as well as a typical large Australian superfund but with higher fees. Since I am forced to contribute to a super currently I think the answer would have to be a yes. 

I might do some further analysis on this (now I have sunk time into extracting the data) when I have a moment. Perhaps look at the impact of GDP shocks on these institutions or perhaps diving into why TNRBF and VNPF have recently been experiencing low returns. 




